Sampling-based Motion Planning for Tracking Evolution of Dynamic Tunnels in Molecular Dynamics Simulations

### 2017-10-31 paper.tex
experiments, discussion, other minor changes in text

\section{Related work}

@@

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling e.g.
around medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting
uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its approaximation~\cite{hollemanMAPRM}.
Another approach is to estimate difficult regions online, e.g. based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
Generating random samples in difficult regions however does not ensure the construction of a better tree in the case of RRT-based methods.
As RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth
of the tree is not blocked by obstacles.
Instead of increasing probability of sampling in all regions at once, e.g. along the whole medial axis, the tree has to be guided through the space to reach the regions gradually.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along
the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.

@@

\section{Algorithm overview}

@@

\section{Used notation}

@@

The main purpose of the tunnel detection is the identification of tunnels, either static or dynamic, that connect the active site with the protein surface.
We assume that the position (configuration) of the active site $\qinit \in \CF^i$ in each snapshot is known in advance 
or it is known that the probe cannot be placed at the active site.
A static tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^i$, i.e., both active site
and the exit point are located in the same frame $i$.
A dynamic tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^j$, $j>i$ i.e., the active size and the
 exit point are located in different frames.

@@

\section{Void space detection in a single frame}

@@

\subsection{Protection from sampling the protein outer space}

@@

To ensure that the tree explores primarily the internal void space while allowing its growth outside the protein for the purpose of exit point detection, the algorithm is divided into two phases.
In the first phase, the growth of the tree outside the protein is blocked using a set of {\sl blocking spheres} $\SB$.
The blocking spheres are placed at configurations that are located close to the surface atoms $\SSA^i$.
After a configuration $q$ is added to the tree, the distance between $q$ and the surface atoms is computed.
If the distance is smaller than a predefined threshold $\dts$, $q$ can be assumed to be close to the surface and a new blocking sphere of radius $\gprobe$ is added to $\SB$.
By adding the blocking sphere at $q$, the tree cannot further expand not only from $q$, but also from the nearby nodes, which ensures,
   that the tree will not exit the protein in the first phase of the void space detection.

@@

\subsection{Voronoi-based guided sampling}
\label{sec::vbg}

@@

\subsection{Expansion of the tree}

@@

\section{Detecting void space through the protein dynamics}

Detection of dynamic tunnels requires to track the evolution of the void space through the molecular dynamics simulation.
The sequence of protein dynamics is searched by the repeated construction of the collision-free tree and its transfer to the next frame.
The relations between the void space of the previous frame and the void space transfered to the new frame are maintained.
It is necessary to support spliting and merging of the detected void space.
The splitting is caused by atoms that move in the next frame to a part that was detected as void space in the previous frame.
This splitting is achieved during the transfer of the tree to the next frame.
The merging is an inverse process, where a previously unconnected regions of the void space can be connected again.
The splitting and merging are illustrated in Fig.~\ref{fig::merging}.
These two procedures are described in the next subsections.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccccc}
\includegraphics[width=0.15\textwidth]{fig/merging1} &
\includegraphics[width=0.15\textwidth]{fig/merging2} &
\includegraphics[width=0.15\textwidth]{fig/merging3} &
\includegraphics[width=0.15\textwidth]{fig/merging4} &
\includegraphics[width=0.15\textwidth]{fig/merging5} &
\includegraphics[width=0.15\textwidth]{fig/merging6}  \\
a (frame $i$) & b (frame $i+1$) & c (frame $i+1$) & d (frame $i+2$)  & e (frame $i+2$) & f (frame $i+2)$ \\
detected      & tree transfer   & void space      & tree transfer    & void space      & merging   \\
void space    & (spliting)      & detection       &  (splitting)     &  detection      & \\              
\end{tabular}
\caption{\label{fig::merging}
An example of merging.
The tree in the first frame with a denoted motion of one of the atoms (a). 
A path exists between the nodes $A$ and $B$.
The moving atom in the second frame causes splitting of the detected void space to three components (b).
The components are expanded by new nodes (green) in the frame $i+1$ (c), but they are still too far so merging cannot connect them.
There is not any path between $A$ and $B$ in the frame $i+1$. 
In the third frame, atoms move back without colliding with any of the void space nodes (d).
The three components are further expanded (green nodes) in the third frame (e), but still there is not any path between $A$ and $B$.
The merging in the $i+2$ frame connects the tree components again (dotted lines) (f).
The result of the merging process is again an existing path between $A$ and $B$.
}
}
\end{figure}

\subsection{Tree transfer}

@@

\subsection{Void space merging}

After the tree  pruning (Alg.~\ref{alg::transfer}), the parents of the surviving nodes point to the previous frame and the edges connecting the nodes are omitted.
The merging is realized by finding nodes in the tree that can be connected by a collisoin-free line.
For each node $q$ of the tree, the $k$-nearest neighbors are found and the possibility of their connection is evaluated using collision
detection on the line between them.
The pseudocode of the void-space merging is listed in Alg.~\ref{alg::component}.

@@

\subsection{Tracking evolution of void space in protein dynamics}

This section presents the main algorithm to track the evolution of the void space in protein dynamics.
This algorithm relies on all the previously described methods and it is listed in Alg.~\ref{alg::dynamic}.
The search starts with the first frame in the configuration $\qinit \in \CF^1$ that represents the initial position of the active site.
The tree from the previous frame is first transferred to the actual frame and the void space is searched in the given frame using Alg.~\ref{alg::main}.
This results in a new configuration tree and also in a new set of exit points.
After $n$ frames are processed, the algorithm results in $n$ configuration trees.

@@

\subsection{Reconstruction of the static and dynamic tunnels}

After the void space is detected in all frames by Alg.~\ref{alg::dynamic}, the tunnels can be identified using the graph $G$ describing
evolution of the void space.
The nodes in this graph are the components of the void space in each frame, the directed edges connect the components from the frame $i$ with the previous frame $i-1$ and the undirected edges connect the exit points and the active sites to their components.

@@

\subsection{Implementation details}

The proposed algorithm relies on two main techniques: collision detection and the nearest-neighbor search.
Fast collision detection can be generally realized using hiearchical data structures like AABB or OBB trees~\cite{ericsonRTCD}.
BioCD collision detection is ~\cite{angulo2005biocd} is dedicated for protein structures.
The nearest-neighbor search can be realized fast using KD-trees. 
The new nodes added to the configuration tree has to be also added to the KD-tree structure, so the KD-tree should support
fast insert operation. 
MPNN~\cite{yershovaMPNN} was designed for the purpose of sampling-based motion planinng and it is used also in our implementation.

@@

In comparison to classic RRT, which adds only one node to the tree per iteration, the proposed method can expand the tree by up to $m+1$ nodes.
Consequently, the number of collision detection queries is $m$ times higher in each iteration.
The nearest-neighbor search is used more times, than in classic RRT, because  the nearest-neighbor search is used to update the active Voronoi vertices.
The complexity of the update of the active Voronoi vertices is $\mathcal{O}(|\VV^i|\log(|\T|))$, where $|\T|$ is the number of nodes in the tree, assuming that the nearest neighbor search is realized using KD-trees.
For the sake of simplicity, we update $\VVA^i$ in each step in the pseudocode (Alg.~\ref{alg::main}), but practically, it is not necessary.
The tree expands incrementally, and usually it takes many iterations before most of the active Voronoi vertices are
reached by the tree.
It is therefore possible to update these vertices e.g. in every $n$-th iteration.

@@

Each frame can be processed separately and it is not necessary to hold all the data structures from all previous frames in memory.
This is valid for all data structures like nearest-neighbor search, collision detection and the configuration tree.
The only exception is the tree pruning (Alg.~\ref{alg::transfer}), where both the tree from the previous frame and the new tree has to be stored in the memory.
After the tree is transfered to the next frame, the old tree does not need to be stored in the memory.
The reconstruction of the tunnels requires to search graph of the compoments. 
As each frame results in few (or maximally tens) of components, hence searching this graph also does not require high amount of memory.

@@

\section{Experimental verification}

A sequence of 8000 frames of protein Haloalkane dehalogenase (DhaA, PDB name 1CQW) was analyzed.
The molecular dynamics was computed using the AMBER tool.
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
The tunnels were investigated for $\probe=1.4$~\AA.

In each frame, static tunnels were detected using CAVER 3.0~\cite{caver3} state-of-the-art tool.
The exits of the tunnels were defined using solvent radius 4~\AA.
In 2860 frames, no static tunnel was found at all, 5100 frames contain one tunnel and in 40 frames, even two tunnels were found.

The dynamic tunnels were detected using the proposed algorithm.
The positions of the active sites were take from the result of the static tunnel detection (from those frames, where at least
one static tunnel was detected).
%(i.e., the position of the active site was not known for 2860 frames).
The goal region of the protein was defined in the same way as for the static tunnel detection, i.e., $\gprobe=4$~\AA\ and 
the surface atoms were detected using $\alpha$-shape with the same radius.

In each frame, the void space detection (Alg.~\ref{alg::main}) was performed with $\Imax=1200$ iterations.
The second phase of the search was switched after $\alpha=70$~\% of the iterations (i.e., after 840 iterations).
Other parameters used for the computations are: $m=8$, $\da=5$~\AA, $\db=0.5$~\AA, $\gb=0.1$ and $\dts=2$~\AA.

In each iteration, up to $m=8$ new samples can be added to the tree which could result in up to 
$1200 \times 8 = 9600$ new nodes in each frame. 
The number of the tree nodes does not rise during the computation, because many of the nodes are pruned.
The average tree size is 3924 (std.dev 1154) nodes.
The size of the built tree together with the number of detected compoments in each frame are depicted in Fig.~\ref{fig::treesize}.
Example of the detected void space is is depicted in Fig.~\ref{fig::dyn}.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{fig/probe14treeSize} &
\includegraphics[width=0.48\textwidth]{fig/probe14components}  \\
a & b  \\
\end{tabular}
}
\caption{\label{fig::treesize}
    The number of nodes in the tree representing the void space (a) and the number of graph components in each frame
        of protein dynamics (b).
}
\end{figure}

\begin{figure*}
\centering
{
\def\ee{0.26}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{cccc}
\includegraphics[width=\ee\textwidth]{fig/video-0560} &
\includegraphics[width=\ee\textwidth]{fig/video-0561} &
\includegraphics[width=\ee\textwidth]{fig/video-0562} &
\includegraphics[width=\ee\textwidth]{fig/video-0563}  \\
Frame 560 & Frame 561 & Frame 562 & Frame 563 \\                       
\includegraphics[width=\ee\textwidth]{fig/video-0699} &
\includegraphics[width=\ee\textwidth]{fig/video-0700} &
\includegraphics[width=\ee\textwidth]{fig/video-0701} &
\includegraphics[width=\ee\textwidth]{fig/video-0702}  \\
Frame 699 & Frame 700 & Frame 701 & Frame 702   \\                       
\end{tabular}
}
\caption{\label{fig::dyn}
Evolution of the void space in molecular dynamics of the DhaA protein. 
In the frames 560--563, active site is presented and it is always connected to the void space (red sphere).
Contrary, no active site was detected (for probe 1.4~\AA) in the frames 699--701.
After it appears again in the frame 702, it is connected to the void space.
}
\end{figure*}

\begin{table}
\centering
\caption{\label{tab::dyn} The number of frames with detected dynamic tunnels. The number of frames
    with static tunnels was detection by CAVER 3.0. $^{*}$ number of static tunnels found by the proposed method.
}
\begin{tabular}{ccccc}
\toprule
                              &   &  \multicolumn{2}{c}{ Static tunnels}   \\
                              &    &  {\bf Exists} & {\bf Not exists} & {\bf Total } \\
\midrule                                     
\multirow{3}{*}{Dynamic tunnels} & {\bf Detected} & 1324 & 1027 & 2351 \\
                                 & {\bf Not detected} & 3816/3118$^*$ & 1833 & 5649 \\
                                 & {\bf Total }       &  5140 & 2860 \\
\bottomrule                                 
\end{tabular}
\end{table}

The number of frames where dynamic tunnels were detected is summarized in Tab.~\ref{tab::dyn}.
In 2860 frames, no static tunnel was found by CAVER, but  in 1027 of these cases, a dynamic tunnel was detected.
A static tunnel and also a dynamic tunnel was detected in 1324 frames.
These results bring new information about the protein, as they show, that a pathway to the active site exists more
frequently than is described by the classic static tunnels.
The examples of frames without a static tunnel, but with a dynamic one, are depicted in Fig.~\ref{fig::seq1}.

In the case of 3816 frames where a static tunnel was detected by CAVER, no dynamic tunnel was detected by the proposed method. 
3118 frames out of 3816 was identified as static.
This indicates, that some tunnels were detected as dynamic instead of static.
The proposed method is non-deterministic and it may happen, that some part of the tree does not
reach the outer region of the protein and consequently, some exit points are not detected.
This decreases the number of detected static tunnels.
These missed exit points are then reached in the next frame which leads to detection of the dynamic tunnel.
The number of correctly classified static tunnels can be increased by using higher $\Imax$.
In our experiment, $\Imax=1200$ was used lower in order to process the whole sequence of 8000 frames in a reasonable time.

The computation was realized on Intel(R) Xeon(R) CPU E5-2630 v2 @ 2.60GHz with 64GB of RAM using one thread.
As the algorithm holds in the memory only the actual tree (which has less than 10,000 nodes in the most of the cases),
the maximally allocated memory during the computation was less than 1GB.
The average computation time to process one frame is 15~s.
In comparison, the computation of the static tunnels using CAVER 3.0 tool takes $\sim$11s per frame.
The processing times for a long sequence of 8000 frames leads to around 33 hours of computation on a single processor.
Such time is still negligible in the comparison to the preparation of the molecular dynamics simulation.
%Times: 8000 frames, cca 35 hodin = cca 15s/frame. Caver na mem ntb: cca 11s/frame.

Many dynamic tunnels were discovered in the studied protein dynamics.
Simple inspection of the detected tunnels consists of their visualization, e.g. by showing all part of the dynamic
tunnel in one frame.
Example of this type of visualization is depicted in Fig.~\ref{fig::joint}.
Such visualization allows chemists to identify surrounding residues and determine possible chemical prosperties
of the tunnels.
For analysis of the large sequence, aggregation of the results is necessary.
In~\cite{fig::vis1}, dynamic tunnels are visualized bar vertical bars colored according to the number of frames occupied by the tunnel.

\begin{figure}
\centering
\includegraphics[width=0.33\textwidth]{fig/58-60crop}
\includegraphics[width=0.33\textwidth]{fig/280-283crop}
\includegraphics[width=0.28\textwidth]{fig/263-267crop}
\caption{\label{fig::joint}
Visualization of dynamic tunnels evolving over three, four and five frames.
Each part is depicted in different collor
}
\end{figure}

\begin{figure*}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/framed-0033-001-0032} & 
\includegraphics[width=0.24\textwidth]{fig/framed-0033-002-0033label} &
\includegraphics[width=0.24\textwidth]{fig/framed-0033-002-0032} & 
\includegraphics[width=0.24\textwidth]{fig/framed-0033-001-0033label} \\
\multicolumn{2}{c}{tunnel I, frames 32-33} &
\multicolumn{2}{c}{tunnel II, frames 32-33} \\
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0049} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0050} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0051} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0053} \\
\multicolumn{4}{c}{tunnel I, frames 49--52} \\
\end{tabular}
\caption{\label{fig::seq1}
Examples of dynamic tunnels.
Two dynamic tunnels were detected in the frames 32-33, both end in different exit point (P1 and P2).
The bottom sequence shows evolution of a tunnel over four frames.
The active site (red sphere) is not presented in frame 33, 50 and 52, which indicates that no static tunnel
was found in these frames.
}
}
\end{figure*}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}

@@

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling, e.g., around medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its approaximation~\cite{hollemanMAPRM}.
Another approach is to estimate difficult regions online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
Generating random samples in difficult regions however does not ensure the construction of a better tree in the case of RRT-based methods.
As RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth of the tree is not blocked by obstacles.
Instead of increasing probability of sampling in all regions at once, e.g., along the whole medial axis, the tree has to be guided through the space to reach the regions gradually.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.

@@

\section{Algorithm Overview}

@@

\section{Used Notation}

@@

The main purpose of the tunnel detection is the identification of tunnels, either static or dynamic, that connect the active site with the protein surface.
We assume that the position (configuration) of the active site $\qinit \in \CF^i$ in each snapshot is known in advance or it is known that the probe cannot be placed at the active site.
A static tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^i$, i.e., both active site and the exit point are located in the same frame $i$.
A dynamic tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^j$, $j>i$, i.e., the active size and the exit point are located in different frames.

@@

\section{Void Space Detection in a Single Frame}

@@

\subsection{Protection from Sampling the Protein Outer Space}

@@

To ensure that the tree explores primarily the internal void space while allowing its growth outside the protein for the purpose of exit point detection, the algorithm is divided into two phases.
In the first phase, the growth of the tree outside the protein is blocked using a set of {\sl blocking spheres} $\SB$.
The blocking spheres are placed at configurations that are located close to the surface atoms $\SSA^i$.
After a configuration $q$ is added to the tree, the distance between $q$ and the surface atoms is computed.
If the distance is smaller than a predefined threshold $\dts$, $q$ can be assumed to be close to the surface and a new blocking sphere of radius $\gprobe$ is added to $\SB$.
By adding the blocking sphere at $q$, the tree cannot further expand not only from $q$, but also from the nearby nodes, which ensures that the tree will not exit the protein in the first phase of the void space detection.

@@

\subsection{Voronoi-Based Guided Sampling}
\label{sec::vbg}

@@

\subsection{Expansion of the Tree}

@@

\section{Detecting Void Space Through the Protein Dynamics}

Detection of dynamic tunnels requires to track the evolution of the void space through the molecular dynamics simulation.
The sequence of protein dynamics is searched by the repeated construction of the collision-free tree and its transfer to the next frame.
The relations between the void space of the previous frame and the void space transferred to the new frame are maintained.
It is necessary to support splitting and merging of the detected void space.
The splitting is caused by atoms that move in the next frame to a part that was detected as void space in the previous frame.
This splitting is achieved during the transfer of the tree to the next frame.
The merging is an inverse process, where a previously unconnected regions of the void space can be connected again.
The splitting and merging are illustrated in Fig.~\ref{fig::merging}.
These two procedures are described in the next subsections.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccccc}
\includegraphics[width=0.15\textwidth]{fig/merging1} &
\includegraphics[width=0.15\textwidth]{fig/merging2} &
\includegraphics[width=0.15\textwidth]{fig/merging3} &
\includegraphics[width=0.15\textwidth]{fig/merging4} &
\includegraphics[width=0.15\textwidth]{fig/merging5} &
\includegraphics[width=0.15\textwidth]{fig/merging6}  \\
a (frame $i$) & b (frame $i+1$) & c (frame $i+1$) & d (frame $i+2$)  & e (frame $i+2$) & f (frame $i+2)$ \\
detected      & tree transfer   & void space      & tree transfer    & void space      & merging   \\
void space    & (splitting)      & detection       &  (splitting)     &  detection      & \\              
\end{tabular}
\caption{\label{fig::merging}
An example of merging.
The tree in the first frame with a denoted motion of one of the atoms (a). 
A path exists between the nodes $A$ and $B$.
The moving atom in the second frame causes splitting of the detected void space to three components (b).
The components are expanded by new nodes (green) in the frame $i+1$ (c), but they are still too far so merging cannot connect them.
There is not any path between $A$ and $B$ in the frame $i+1$. 
In the third frame, atoms move back without colliding with any of the void space nodes (d).
The three components are further expanded (green nodes) in the third frame (e), but still there is not any path between $A$ and $B$.
The merging in the $i+2$ frame connects the tree components again (dotted lines) (f).
The result of the merging process is again the existing path between $A$ and $B$.
}
}
\end{figure}

\subsection{Tree Transfer}

@@

\subsection{Void Space Merging}

After the tree pruning (Alg.~\ref{alg::transfer}), the parents of the surviving nodes point to the previous frame and the edges connecting the nodes are omitted.
The merging is realized by finding nodes in the tree that can be connected by a collision-free line.
For each node $q$ of the tree, the $k$-nearest neighbors are found and the possibility of their connection is evaluated using collision
detection on the line between them.
The pseudocode of the void-space merging is listed in Alg.~\ref{alg::component}.

@@

\subsection{Tracking the Evolution of Void Space in Molecular Dynamics}

This section presents the main algorithm to track the evolution of the void space in molecular dynamics.
This algorithm relies on all the previously described methods and it is listed in Alg.~\ref{alg::dynamic}.
The search starts with the first frame in the configuration $\qinit \in \CF^1$ that represents the initial position of the active site.
The tree from the previous frame is first transferred to the actual frame and the void space is searched in the given frame using Alg.~\ref{alg::main}.
This results in a new configuration tree and also in a new set of exit points.
After $n$ frames are processed, the algorithm results in $n$ configuration trees.

@@

\subsection{Reconstruction of the Static and Dynamic Tunnels}

After the void space is detected in all frames by Alg.~\ref{alg::dynamic}, the tunnels can be identified using the graph $G$ describing the evolution of the void space.
The nodes in this graph are the components of the void space in each frame, the directed edges connect the components from the frame $i$ with the previous frame $i-1$ and the undirected edges connect the exit points and the active sites to their components.

@@

\subsection{Implementation Details}

The proposed algorithm relies on two main techniques: the collision detection and the nearest-neighbor search.
Fast collision detection can be generally realized using hierarchical data structures like AABB or OBB trees~\cite{ericsonRTCD}.
The BioCD collision detection is ~\cite{angulo2005biocd} is dedicated for protein structures.
The nearest-neighbor search can be realized fast using KD-trees. 
The new nodes added to the configuration tree has to be also added to the KD-tree structure, so the KD-tree should support
fast insert operation. 
MPNN~\cite{yershovaMPNN} was designed for the purpose of sampling-based motion planning and it is used also in our implementation.

@@

In comparison to classic RRT, which adds only one node to the tree per iteration, the proposed method can expand the tree by up to $m+1$ nodes.
Consequently, the number of collision detection queries is $m$ times higher in each iteration.
The nearest-neighbor search is used more times, than in classic RRT, because  the nearest-neighbor search is used to update the active Voronoi vertices.
The complexity of the update of the active Voronoi vertices is $\mathcal{O}(|\VV^i|\log(|\T|))$, where $|\T|$ is the number of nodes in the tree, assuming that the nearest neighbor search is realized using KD-trees.
For the sake of simplicity, we update $\VVA^i$ in each step in the pseudocode (Alg.~\ref{alg::main}), but practically, it is not necessary.
The tree expands incrementally, and usually it takes many iterations before most of the active Voronoi vertices are reached by the tree.
It is therefore possible to update these vertices, e.g., in every $n$-th iteration.

@@

Each frame can be processed separately and it is not necessary to hold all the data structures from all previous frames in memory.
This is valid for all data structures like nearest-neighbor search, collision detection and the configuration tree.
The only exception is the tree pruning (Alg.~\ref{alg::transfer}), where both the tree from the previous frame and the new tree has to be stored in the memory.
After the tree is transferred to the next frame, the old tree does not need to be stored in the memory.
The reconstruction of the tunnels requires to search graph of the components. 
As each frame results in few (or maximally tens) of components, hence searching this graph also does not have high memory requirements.

@@

\section{Experimental Verification}

Our proposed algorithm has been tested on several MD simulations obtained by the domain experts. 
Here we present the results of one such a study, performed on an MD consisting of 8,000 frames of the DhaA haloalkane dehalogenase protein (PDB ID 1CQW).
The molecular dynamics was simulated using the AMBER tool.
The active site was defined by the surrounding residues with sequence numbers 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
The tunnels were investigated for $\probe=1.4$~\AA.

First, in each frame the static tunnels were detected using the CAVER 3.0~\cite{caver3} state-of-the-art tool.
The exits of these tunnels were defined using solvent radius 4~\AA.
In 2,860 frames, the CAVER tool did not detect any static tunnel, 5,100 frames contained only one tunnel, and in 40 frames CAVER found even two tunnels.

The same dataset has been scrutinized by our proposed algorithm for detection of dynamic tunnels.
The positions of the active site in each frame were taken from the results of the static tunnel detection (from those frames, where at least
one static tunnel was detected).
%(i.e., the position of the active site was not known for 2860 frames).
The goal region of the protein was defined in the same way as for the static tunnel detection, i.e., $\gprobe=4$~\AA\ and 
the surface atoms were detected using the $\alpha$-shape with the same radius.

In each frame, the void space detection (Alg.~\ref{alg::main}) was performed with $\Imax=1200$ iterations.
The second phase of the search was switched after $\alpha=70$~\% of the iterations (i.e., after 840 iterations).
Other parameters and their values used for the computation were $m=8$, $\da=5$~\AA, $\db=0.5$~\AA, $\gb=0.1$, and $\dts=2$~\AA.

In each iteration, up to $m=8$ new samples can be added to the tree which can result in up to $1,200 \times 8 = 9,600$ new nodes in each frame. 
The number of the tree nodes does not rise during the computation because many of the nodes are pruned.
The average tree size is 3,924 (std.dev 1,154) nodes.
The size of the built tree together with the number of detected components in each frame are depicted in Fig.~\ref{fig::treesize}.
The example of the detected void space is depicted in Fig.~\ref{fig::dyn}.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{fig/probe14treeSize} &
\includegraphics[width=0.48\textwidth]{fig/probe14components}  \\
a & b  \\
\end{tabular}
}
\caption{\label{fig::treesize}
    The number of nodes in the tree representing the void space (a) and the number of graph components in each frame
        of molecular dynamics simulation (b).
}
\end{figure}

\begin{figure*}
\centering
{
\def\ee{0.26}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{cccc}
\includegraphics[width=\ee\textwidth]{fig/video-0560} &
\includegraphics[width=\ee\textwidth]{fig/video-0561} &
\includegraphics[width=\ee\textwidth]{fig/video-0562} &
\includegraphics[width=\ee\textwidth]{fig/video-0563}  \\
Frame 560 & Frame 561 & Frame 562 & Frame 563 \\                       
\includegraphics[width=\ee\textwidth]{fig/video-0699} &
\includegraphics[width=\ee\textwidth]{fig/video-0700} &
\includegraphics[width=\ee\textwidth]{fig/video-0701} &
\includegraphics[width=\ee\textwidth]{fig/video-0702}  \\
Frame 699 & Frame 700 & Frame 701 & Frame 702   \\                       
\end{tabular}
}
\caption{\label{fig::dyn}
Evolution of the void space in molecular dynamics of the DhaA protein. 
In frames 560--563, the active site is contained in the detected void space (red sphere).
On the contrary, the void space (for probe 1.4~\AA) detected in frames 699--701 does not contain the active site.
The active site appears again inside the void space in frame 702.
}
\end{figure*}

\begin{table}
\centering
\caption{\label{tab::dyn} The number of frames with detected dynamic tunnels. The number of frames
    with static tunnels, detected by CAVER 3.0. $^{*}$ number of static tunnels found by the proposed method.
}
\begin{tabular}{ccccc}
\toprule
                              &   &  \multicolumn{2}{c}{ Static tunnels}   \\
                              &    &  {\bf Exists} & {\bf Does not exist} & {\bf Total } \\
\midrule                                     
\multirow{3}{*}{Dynamic tunnels} & {\bf Detected} & 1,324 & 1,027 & 2,351 \\
                                 & {\bf Not detected} & 3,816/3,118$^*$ & 1,833 & 5,649 \\
                                 & {\bf Total }       &  5,140 & 2,860 \\
\bottomrule                                 
\end{tabular}
\end{table}

The number of frames where the dynamic tunnels were detected is summarized in Tab.~\ref{tab::dyn}.
In 2,860 frames, no static tunnel was found by CAVER, but in 1,027 frames of this set, a dynamic tunnel was detected.
Both static and dynamic tunnels were detected in 1,324 frames.
These results bring new information about the protein behavior. 
They show that a path connecting the protein surface with the active site is more frequent than the static tunnels suggest.
The examples of frames without a static tunnel, but containing a dynamic one, are depicted in Fig.~\ref{fig::seq1}.

In the case of 3,816 frames where a static tunnel was detected by CAVER, no dynamic tunnel was detected by the proposed method. 
3,118 frames out of 3,816 were identified as static.
This indicates that some tunnels were detected as dynamic instead of static.
The proposed method is non-deterministic and it may happen, that some part of the tree does not reach the outer region of the protein and consequently, some exit points are not detected.
This decreases the number of detected static tunnels.
These missed exit points are then reached in the next frame which leads to the detection of the dynamic tunnel.
The number of correctly classified static tunnels can be increased by using higher $\Imax$.
In our experiment, $\Imax=1,200$ was used with a lower value in order to process the whole sequence of 8,000 frames in a reasonable time.

The computation was realized on Intel(R) Xeon(R) CPU E5-2630 v2 @ 2.60GHz with 64GB of RAM using one thread.
As the algorithm holds in the memory only the actual tree (which has less than 10,000 nodes in most of the cases),
the maximally allocated memory during the computation was less than 1GB.
The average computation time to process one frame is 15~s.
In comparison, the computation of the static tunnels using CAVER 3.0 tool takes $\sim$11s per frame.
The processing times for a long sequence of 8,000 frames leads to around 33 hours of computation on a single processor.
Such time is still negligible in the comparison to the preparation of the molecular dynamics simulation.
%Times: 8000 frames, cca 35 hodin = cca 15s/frame. Caver na mem ntb: cca 11s/frame.

Many dynamic tunnels were discovered in the studied protein dynamics.
The inspection of the detected tunnels utilizes a simple visualization which shows all parts of the dynamic tunnel in one frame.
The example of this type of visualization is depicted in Fig.~\ref{fig::joint}.
Such visualization allows the biochemists to identify the surrounding residues and further study their physico-chemical properties and their influence on the tunnels.
For the analysis of a large simulation, an aggregation of the results is necessary.
In~\cite{fig::vis1}, the dynamic tunnels are visualized using vertical bars colored according to the number of frames occupied by this tunnel.

\begin{figure}
\centering
\includegraphics[width=0.33\textwidth]{fig/58-60crop}
\includegraphics[width=0.33\textwidth]{fig/280-283crop}
\includegraphics[width=0.28\textwidth]{fig/263-267crop}
\caption{\label{fig::joint}
Examples of visualization of the dynamic tunnels evolving over three, four, and five frames. Different colors depict different frames of the dynamic tunnel. This enables the users to easily determine the portion (length) of the tunnel which was detected in the corresponding frame. 
}
\end{figure}

\begin{figure*}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/framed-0033-001-0032} & 
\includegraphics[width=0.24\textwidth]{fig/framed-0033-002-0033label} &
\includegraphics[width=0.24\textwidth]{fig/framed-0033-002-0032} & 
\includegraphics[width=0.24\textwidth]{fig/framed-0033-001-0033label} \\
\multicolumn{2}{c}{tunnel I, frames 32-33} &
\multicolumn{2}{c}{tunnel II, frames 32-33} \\
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0049} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0050} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0051} &
\includegraphics[width=0.24\textwidth]{fig/framed-0053-000-0053} \\
\multicolumn{4}{c}{tunnel I, frames 49--52} \\
\end{tabular}
\caption{\label{fig::seq1}
Examples of dynamic tunnels.
Two dynamic tunnels were detected in frames 32-33, both ended in different exit points (P1 and P2).
The bottom sequence shows the evolution of a dynamic tunnel over four frames.
The active site (red sphere) is not inside the detected void space in frames 33, 50, and 52, which indicates that any static tunnels
were detected in these frames.
}
}
\end{figure*}

\begin{figure*}
\centering
%\includegraphics[width=0.99\textwidth]{fig/dyn_tunnels_7985}
\includegraphics[width=0.99\textwidth]{fig/5}
\caption{\label{fig::vis1}
Visualization showing the results of the dynamic tunnel detection performed on the MD simulation containing 8,000 frames. 
  Each line corresponds to one detected dynamic tunnel.
  The length of the line depicts the actual length of the tunnel (its centerline). 
  The color corresponds to the number of consecutive frames in whose the given dynamic tunnel was detected (i.e., within these frames the tunnel was able to connect the protein surface with its active site).
    It is clearly visible that most of the dynamic tunnels were constructed within two or three frames. However, there are several tunnels spanning through more frames (up to 35).
}
\end{figure*}

{\textcolor{red}{TODO Fig.~\ref{fig::caver34}.}}
